Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-05-31 ThesisJP/Introduction/introduction.tex
Finished all comments from Sverre on the english text.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.

This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed. Algorithms that previously had to be implemented in hardware, for performance reasons, can now be implemented in software. 
%Algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, become realizable. 
It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.

The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing to happen in near future. To achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second. Nevertheless, with modern GPUs, this amount of processing is finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging, it is important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Paper\,I} and \textbf{II}]. However, two additional methods have also been explored. The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the second one being the simulation of dense ultrasound pressure fields [\textbf{Paper\,V}]. Both share the property of being computationally complex, but on the other hand they consist of many independent computations, which makes them perfectly suited for parallel GPU acceleration.

@@

\subsection{Paper\,I}
\textbf{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the presented implementation is applicable to a range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a less strict real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing.

@@

\subsection{Paper\,II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something that has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging over the past decade. In \textbf{Paper\,I}, arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming in cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

In this paper, we implement the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}) on the GPU  to reduce the covariance matrix size. For parameters that give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64- and 96-element linear arrays.  \textbf{Paper\,II} and \textbf{VI} also present, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

@@

Further, it has been investigated how the Capon beamforming achieves its super-resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector. Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves across an imaging sector which has insufficient lateral sampling. With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.  To obtain the same lateral gain variations as for \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required. The same gain variations are observed for broadband nearfield simulations. Finally, we show that the gain variation can be reduced by oversampling on receive using phase rotation steering, hence improving shift-invariance. This is achieved without affecting the acquisition frame rate and with only a minor increase in computational complexity. The method is successfully applied on simulations and \textit{in vitro} phantom data.

\subsection{Paper\,IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
The fourth paper proposes an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization methods to an ultrasound imaging system and still have real-time performance.

@@

The work in \textbf{Paper\,II} made it clear that interesting findings are done when a real-time implementation makes it possible to watch the temporal behavior of an algorithm. An observation of temporal artifacts encourage for the work of \textbf{Paper\,III}.

>>>>

\section{Motivation}
The rapid development of computer game technology and accompanying programming languages, as shown in Chapter \ref{chap1}, have recently provided researchers with small personal supercomputers, comprised in a single graphics processing unit (\nom{GPU}{Graphics processing unit}). %The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.
This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed. Algorithms that previously had to be implemented in hardware, for performance reasons, can now be implemented in software. 
%Algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, become realizable. 
It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.

The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing to happen in near future. As shown in Section \ref{sec:adaptbf}, an effective processing rate of hundreds of GFLOPS is required to achieve real-time processing for cardiac ultrasound imaging. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second. Nevertheless, with modern GPUs, this amount of processing is finally available within a single GPU card. When researchers are exploring new algorithms for ultrasound imaging, it is important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Papers\,I} and \textbf{II}]. However, two additional methods have also been explored. The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the second one being the simulation of dense ultrasound pressure fields [\textbf{Paper\,V}]. Both share the property of being computationally complex, but on the other hand they consist of many independent computations, which makes them perfectly suited for parallel GPU acceleration.

@@

\subsection{Paper\,I}
\textbf{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the presented implementation is applicable to a range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a less strict real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing. The transmit beams used for sonar imaging are also much wider than the once used for ultrasound imaging. This makes techniques like the BS-Capon beamformer less attractive for sonar imaging. However, as shown in this paper, low-complexity methods are not needed to reach real time Capon beamforming for sonar imaging.

@@

\subsection{Paper\,II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance in the future. This is something that has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging over the past decade. In \textbf{Paper\,I}, arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming in cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

In this paper, we implement the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}) on the GPU  to reduce the covariance matrix size. For parameters that give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64- and 96-element linear arrays.  \textbf{Papers\,II} and \textbf{VI} also present, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

@@

Further, it has been investigated how the Capon beamforming achieves its super-resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector. Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects move across an imaging sector which has insufficient lateral sampling. With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.  To obtain the same lateral gain variations as for \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required. The same gain variations are observed for broadband nearfield simulations. Finally, we show that the gain variation can be reduced by oversampling on receive using phase rotation steering, hence improving shift-invariance. This is achieved without affecting the acquisition frame rate and with only a minor increase in computational complexity. The method is successfully applied on simulations and \textit{in vitro} phantom data.

\subsection{Paper\,IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
The fourth paper proposes an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This is because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization methods to an ultrasound imaging system and still have real-time performance.

@@

The work in \textbf{Paper\,II} made it clear that interesting findings are done when a real-time implementation makes it possible to watch the temporal behavior of an algorithm. An observation of temporal artifacts was the fact that stimulated for the work of \textbf{Paper\,III}.
